---
title: "Create Interactive Data Visualization with Plotly in R"
authors:
- admin
- Sophia-Shalhout
date: "2/06/2020"
output:
  word_document: default
  pdf_document: default
  html_document: default
always_allow_html: true
tags: ["Plotly", "Data Science"]
lastMod: "2020-2-06T00:00:00Z"
projects:
  - Data Science
summary: This working document is a user guide on how to create interactive graphs with R via a package called `plotly`
image:
  caption: 'Photo by Carlos Muza on Unsplash'
  focal_point: ""
  placement: 2
  preview_only: false
social:
- icon: twitter
  icon_pack: fab
  link: https://twitter.com/TheMillerLab
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
# Overview
- This is an RMarkdown explaining how we use the `Plotly` package to more effectivley communicate clinical and translational research via interactive graphs
- Plotly is a computing company that develops online data analytics and visualization tools
- It has open sourced many useful interactive visualization products.
- Plotly can be used in several programming languages (e.g. `Python`, `R` and `JavaScript`)
  - This tutorial will be written with instruction for `R` users
  - Although https://plot.ly/ is fabulous resource, we found that there were elements from those tutorials that were missing and we hope that this post will provide a useful resource to get started in `Plotly` in `R`
- Why use Plotly?
  - Plotly allows us to create visually appealing interactive plots. 
  - The ability to export to `html` and retain all interactive functionality is easily accomplished with Plotly.
  - It is focused around the ability to generate interactive plots with a few lines of code.
  - Finally, all interactive features are compatible with modern web browsers.
  
# Step 1, Download Plotly from CRAN
- Use the install.package() function to install the plotly R package from [CRAN](https://cran.r-project.org/)
`install.packages("plotly")`

# Load Relevant Packages
```{r Load Packages, warning=FALSE, message=FALSE}
library(tidyverse)
library(knitr)
library(plotly)
library(readxl)
library(scales)
```


# Load dataset 
- For the first example we will use mtcars (Motor Trend Car Road Tests), which is built into base `R`

```{r mt cars}
mtcars <- mtcars
```

## View the `mtcars` Data Frame
```{r view mtcars}
mtcars %>% kable
```

# Make an Interactive Bar Chart in `Plotly`
- For this first example, we will make a bar chart of the number of vehicles with <font color='red'>"4"</font>, <font color='red'>"6"</font> and <font color='red'>"8"</font> cylinders  
- One approach is to create a new data frame with two vectors (columns)  
  - one that we create with the String "Four Cylinders", "Six Cylinders", and "Eight Cylinders" and  
  - another Vector that is the sum of all of the rows in mtcars that have a <font color='red'>"4"</font>, <font color='red'>"6"</font> or <font color='red'>"8"</font> in them. 
  - We'll call the first vector **"vehicles"** and the second **"cylinders"**  
- This allows us to make a table that has a very defined x-axis and y-axis to make construtcting a bar graph very straight forward
```{r}
vehicles <- c("Four Cylinders","Six Cylinders","Eight Cylinders")
cylinders <- c(sum(mtcars$cyl==4), sum(mtcars$cyl==6), sum(mtcars$cyl==8)) # of note, the sum() function will allow us to add up all the observations with either a "4" or a "6" or a "8" 
```

## View these two vectors
```{r}
vehicles %>% kable
cylinders %>% kable
```

- As you can see, there are 11 vehicles with <font color='red'>"4"</font> contained in the columns' cells, 7 with <font color = 'red'>"6"</font> and 14 with <font color = 'red'>"8"</font>  
- Now combine these two vectors into a tibble, we'll call it <font color = 'green'>"veh_cyl"</font>
  - Of note, a `tibble` is a modern rework of the standard `data.frame`, with some internal improvements to make code more reliable. 
    - They are data frames, but do not follow all of the same rules  
      - For example, tibbles can have column names that are not normally allowed, such as numbers/symbols.
```{r}
veh_cyl <- tibble(vehicles, cylinders)
```

# View new Tibble
```{r}
veh_cyl %>% kable
```

##### Now that we have a tibble that can be easily turned into a bar graph, let's use Plotly to make an interactive graph
```{r}
plot_ly(data = veh_cyl, x = vehicles, y = cylinders, type = "bar", text = cylinders, textposition = "auto") %>% 
  layout(title = "Number of Vehicles in mtcars with 4, 6, and 8 Cylinders",
         titlefont = list(size = 28, color = "orange", family = "Calibri"),
    yaxis = list(title = "Number of Vehicles",
                 titlefont = list(color = "black", family = "Arial", size = 26),
                 tickfont = list(color = "black", family = "Arial", size = 20)),
    xaxis = list(title = "Number of Cylinders",
                 titlefont = list(color = "red", family = "Times New Roman", size = 22),
                 tickfont = list(color = "green", family = "Cambria", size = 18)))%>% 
  layout(margin = list( 
                l = 10,
              r = 10,
              b = 0,
              t = 40)) # Use the layout(margin) function to adjust the margins of the graph
```  

##### Comments about code for Titles and Axes  
- To adjust the title, axes and margins, `plotly` uses different code that perhaps the more popular `ggpolot` package that `R` users are often very familiar with  
  - The `layout()` and `list()` functions perform much of the work for these tasks, as seen above
  - In order to highlight how these are used, we've made the title <font color = 'orange'>Orange</font> and the axis font <font color = 'green'>Green</font> and the axis title <font color = 'red'>Red</font> to emphaise those elements of the code.  
- To add the values of the bar on the top of the bars use `textposition = "auto"` in the first ()  
- You don't have to use those codes to adjust your Titles and Axes, but if you don't a very basic plot will look like this...
    
```{r bar chart basic}
plot_ly(data = veh_cyl, x = vehicles, y = cylinders, type = "bar", text = cylinders)
```
        
# Make a Time Series in Plolty
- We will use a data set downloaded from GitHub that has dates and corresponding data
- This will also highlight how to download data directly from GitHub into your R Studio
- you will need the packages `httr` and `RCurl`

```{r, message=FALSE, warning=FALSE}
library(httr)
library(RCurl)
```

### Download List of publications from Github  
- We are going to uses a dataset from **Reproducible Research in Ecology, Evolution, Behaviour, and Environmental Studies** https://github.com/opetchey/RREEBES/tree/master/Beninca_etal_2008_Nature

```{r}
no2 <- read.csv(text=getURL("https://raw.githubusercontent.com/opetchey/RREEBES/master/Beninca_etal_2008_Nature/data/nutrients_original.csv"), skip=7, header=T)
```

### Convert first column into dates
```{r}
no2$Date <- as.Date(no2$Date, "%d/%m/%y") # of note, this %d/%m/%y format is critical, and is specific to the way the data is put into the data frame; b/c here the date in the original df is d/m/2 digit year, you need a lower case "y" to indicate that year is only two digits in the data frame
```

### Graph a Timeseries in `Plotly`
```{r}
plot_ly( data = no2, x = no2$Date, y = no2$NO2) %>% 
  add_trace(type = "scatter" ,mode = "lines+markers")  %>% 
  layout(
    title = "Time Series of NO2",
    xaxis = list(
      title = "Year"),
    yaxis = list(
      title = "NO2"))
```


# Make a Chloropleth Map in `Plotly`  
- A Chloropleth Map is a map that uses differences in shading, coloring, or the placing of symbols within predefined areas to indicate the average values of a property or quantity in those areas (https://en.wikipedia.org/wiki/Choropleth_map)
- For example let's look at the rate of new cancers in the US according to the 
**Centers for Disease Control and Prevention (CDC) and the National Cancer Institute (NCI), 2016**
- Visit “https://gis.cdc.gov/Cancer/USCS/DataViz.html” and click “export” to download a csv file.  
  - Make sure to save it in your project folder
  - Also, please create a column,”state”, and enter the state abbreviations next to the corresponding full state name in “Area”.  
    - This is needed to add values to each state
    
```{r, }
df <- read.csv("USCS_OverviewMap.csv")
```
### View the Data Fram `df`
```{r view Rate of Cancers in the US according to CDC and NCI 2016 }
df %>% kable
```

### Add a vector to the Data Frame specifying what will be revealed via hovering
```{r hover }
df$hover <- with(df, paste(df$Area, df$Year,"Sex:", df$Sex, "Case count:", df$CaseCount, sep = "<br>"))
```
- This is a very nice option.  
  - You can create a vector (here we are calling it "hover") that will be the contents displayed when you hover over the map  
    - This "hover" vector needs to be specified in the `text =` part of the code below 

#### View the first 2 rows of the data frame `df` with `hover` vector
```{r}
kable(head(df[1:2,]))
```

### Plot the Chloropleth Map
```{r, map of Rate of Cancers in the US, 2016 }
# give state boundaries a white border
l <- list(color = toRGB("white"), width = 2)
# specify some map projection/options
g <- list(
  scope = 'usa',
  projection = list(type = 'albers usa'),
  showlakes = TRUE,
  lakecolor = toRGB('steelblue')
)

p <- plot_geo(df, locationmode = 'USA-states') %>%
  add_trace(
    z = ~AgeAdjustedRate, text = ~hover, locations = df$Code,
    color = ~AgeAdjustedRate, colors = 'Purples'
  ) %>%
  colorbar(title = "Rate per 100,000 people") %>%
  layout(
    title = 'US Cancer Statistics Rate of New Cancers- 2016 (Source: CDC and NCI)',
    geo = g
  )

p
```



# SessionInfo
```{r}
sessionInfo()
```




