---
title: "Graphical Representation of Survival - Swimmer Plots for Clinical Trials in Clinical Oncology"
authors:
- Sophia-Shalhout
- admin
date: "2/09/2020" 
output:
  word_document: default
  pdf_document: default
  html_document: default
always_allow_html: true
tags: ["Swimmer Plots", "Data Science"]
lastMod: "2020-02-09T00:00:00Z"
projects:
  - Data Science
summary: This working document is a user guide on how to create Swimmer Plots to visualize survival data
image:
  caption: 'Photo by Alex Guillaume on Unsplash'
  focal_point: ""
  placement: 2
  preview_only: false
social:
- icon: twitter
  icon_pack: fab
  link: https://twitter.com/TheMillerLab
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
# Overview
- This is an RMarkdown explaining how we can use **<font color='red'>"Swimmer Plots"</font>** to visualize survival data.
- A **<font color='red'>"Swimmer Plot"</font>** is a graphical way of displaying several aspects of a subject’s tumor response such as **total time to tumor response**, whether there was a <font color='green'>"Complete"</font> or <font color='blue'>"Partial"</font> response, and duration of response.
- This is a clear, graphical representation of the course of a patient's tumor response and can be an especially useful tool when reporting clinical trial data results.

# Prepare the Data  

## Step 1, Download ggplot2,  reshape2, dplyr, plotly, and grid from CRAN
- Use the `install.package()` function to install the followng R packages from CRAN: `ggplot2`, `plotly`, `reshape2`, `dplyr`, `kintr` and `grid` from [CRAN](https://cran.r-project.org/) for example:  
`install.packages("ggplot2")`

## Step 2, Load each relevant Package
```{r Load Packages, warning=FALSE, message=FALSE}
library(ggplot2) 
library(dplyr, warn.conflicts=FALSE)   # Useful for manipulating the dataframes
library(reshape2) # Reformmating dataframes
library(grid)
library(plotly) # Allows us to make the swimmer plot interactive
library(knitr)

```

## Step 3, Create an "example" data set for demonstrative purposes
- We will create a working data set appropriate for this type of graphical represenataion.

```{r Create an example data set}
set.seed(35) # This sets the seed of R's random number generator
dat <-  data.frame(Subject = 1:15, 
                 Months = sample(5:20, 15, replace=TRUE), # This generates a random set of months from 5 - 20
                 Treated=sample(0:1, 15, replace=TRUE), # This generates 15 random 0 or 1s which correspond to Tx or no Tx
                 Stage = sample(1:4, 15, replace=TRUE), # This randomly generates staging from 1 - 4 
                 Continued=sample(0:15, 15, replace=TRUE))
```

### View initial Data Set
```{r View dat}
dat %>% kable
```

## Add Response Data to Data Set
```{r Add response data}
dat <-  dat %>%
  group_by(Subject) %>%
  mutate(Complete=sample(c(4:(max(Months)-1),NA), 1, 
                         prob=c(rep(1, length(4:(max(Months)-1))),5), replace=TRUE),
         Partial=sample(c(4:(max(Months)-1),NA), 1, 
                        prob=c(rep(1, length(4:(max(Months)-1))),5), replace=TRUE),
         Durable=sample(c(-0.5,NA), 1, replace=TRUE))
# of note, `sample()`takes a sample of the specified size from the elements of x using either with or without replacement
# Let's organize the order of the Subjects by Months
dat$Subject <-  factor(dat$Subject, levels=dat$Subject[order(dat$Months)])
```

### Let's view the Data Set Now
```{r view data set 2}
dat %>% kable
```

## Melt part of data frame for adding points to bars  
- This will collapse the Columns "Complete", "Partial" and "Durable" into a new column called "variable" and the values of those orginial columns will become a new vector/column called "value"
```{r}
dat.m <-  melt(dat %>% select(Subject, Months, Complete, Partial, Durable),
             id.var=c("Subject","Months"), na.rm = TRUE) 
# of note, na.rm = TRUE will eliminate those rows with missing values
```

### Let's View our Data Set after melting
```{r View Melted Data set}
dat.m %>% kable
```

# Graph the Data using a Swimmer Plot

## Let's make a static swimmer plot with `ggplot`
```{r Graph the Survival Swimmer Plot with ggplot, warning=FALSE}
a<- ggplot(dat, aes(Subject, Months)) +
  geom_bar(stat="identity", aes(fill=factor(Stage)), width=0.8) +
  geom_point(data=dat.m, 
             aes(Subject, value, colour=variable, shape=variable), size=5) +
  geom_segment(data=dat %>% filter(Continued==1), 
             aes(x=Subject, xend=Subject, y=Months + 0.1, yend=Months + 1), 
             pch=15, size=0.8, arrow=arrow(type="closed", length=unit(0.1,"in"))) +
  coord_flip() +
  scale_fill_manual(values=hcl(seq(15,375,length.out=5)[1:4],100,75)) +
  scale_colour_manual(values=c(hcl(seq(15,375,length.out=3)[1:2],100,30),"black")) +
  scale_y_continuous(limits=c(-1,20), breaks=0:20) +
  labs(fill="Disease Stage", colour="", shape="",
       x="Subjects on Treatment Drug A") +
  theme_bw() +
  theme(panel.grid.minor=element_blank(),
        panel.grid.major=element_blank(),
        axis.text.y=element_blank(),
        axis.ticks.y=element_blank())
a

```

## Now let's make an Interactive Swimmer plot in `Plotly` by simply using the `ggplotly()` function of the static plot as an object
```{r, warning=FALSE}
ggplotly(a)
```



# SessionInfo
```{r}
sessionInfo()
```




